---
title: "Skills Problem Set VI"
author: "Earnest Salgado"
date: "25/05/2021"
output: 
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
urlcolor: blue
always_allow_html: true
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

<!-- .Rmd files use  markdown, a text mark up language, to provide formating.--> 
<!--Text include within these strange arrows are comments and will not show up when you knit-->

This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **ES**

Add your collaborators: **GATW**

Late coins used this pset: 0. Late coins left: 6. 
<!--You may use up to two for a given assignment.)-->

# Functions (15 points)
<!--(Notice the use of two `##`)-->

1. Write a function that transforms a vector c("a", "b", "c") into a string, a, b, and c (including the and!!). Think carefully about what the function should do if given a vector of length 1, 2, or 3.

```{r}
str_commasep <- function(x, delim = ",") {
  n <- length(x)
  if (n == 0) {
    ""
  } else if (n == 1) {
    x
  } else if (n == 2) {
    # no comma before and when n == 2
    str_c(x[[1]], "and", x[[2]], sep = " ")
  } else {
    # commas after all n - 1 elements
    not_last <- str_c(x[seq_len(n - 1)], delim)
    # prepend "and" to the last element
    last <- str_c("and", x[[n]], sep = " ")
    # combine parts with spaces
    str_c(c(not_last, last), collapse = " ")
  }
}
str_commasep("")
#> [1] ""
str_commasep("a")
#> [1] "a"
str_commasep(c("a", "b"))
#> [1] "a and b"
str_commasep(c("a", "b", "c"))
#> [1] "a, b, and c"
str_commasep(c("a", "b", "c", "d"))
```

2. Write a function that given your birthday (as a date), returns how old you are in years

```{r}
age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1991-03-17"))
```

3. Statistical functions

  a. Write a function to calculate the variance of a numeric vector
```{r}
x <- c(2, 7, 7, 4, 5, 1, 3)
variance = function(x) {variance = var(x) }

variance <- function(x, na.rm = TRUE) {
  n <- length(x)
  m <- mean(x, na.rm = TRUE)
  sq_err <- (x - m)^2
  sum(sq_err) / (n - 1)
}
```

  b. Write a function to calculate the skewness of a numeric vector
```{r}
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  v <- var(x, na.rm = na.rm)
  (sum((x - m) ^ 3) / (n - 2)) / v ^ (3 / 2)
}

skewness(c(1, 2, 5, 100))
```

  c. Use summarize_if() to calculate the mean, variance, and skewness of all numeric columns in the diamond dataset. Then, tidy the table so we have one row for each variable. (Hint: summarize_if() takes a boolean in the first position and a named vector of functions in the second position.)

```{r}
diamonds  %>%
  summarise_if(is.numeric, funs(Xbar=mean, SD=sd, skew = skewness, var = variance))

```

4. Rename the following functions to better reflect their purpose

```{r}
f <- function(string, prefix){
  str_sub(string, 1, nchar(prefix)) == prefix
}
g <- function(x){
  if (length(x) <= 1)return(NULL)
  x[-length(x)]
}
```

5. Write a greeting function that says "good morning", "good afternoon", or "good evening", depending on the time of day. (Hint: use a time argument that defaults to lubridate::now(). That will make it easier to test your function.)

```{r}
library(lubridate)
greetings=function(time=lubridate::now())
  {#fill-in} greetings()greetings(as.POSIXct("2011-03-27 01:30:00"))
}


greeting <- function(time_now = lubridate::now()) {
  hour_now <- lubridate::hour(time_now)
  
  if (hour_now < 12) {
    "good morning"
  } else if (hour_now < 18) {
    "good afternoon"
  } else {
    "good night"
  }
}
greeting()

```

# For Loops (35 points)

1. Write for loops to
  a. Compute the mean of every column in mtcars
```{r}
output <- vector("double", ncol(mtcars))
    names(output) <- names(mtcars)
    for (i in names(mtcars)) {
      output[i] <- mean(mtcars[[i]])
    }
    output
```

  b. Compute the number of unique values in each column of mpg
```{r}
 data("mpg")
    iris_uniq <- vector("double", ncol(mpg))
    names(mpg_uniq) <- names(mpg)
    for (i in names(mpg)) {
      mpg_uniq[i] <- n_distinct(mpg[[i]])
    }
    mpg_uniq
```

  c. Generate 10 random points distributed poissons (rpois) for each $\lambda$ = 1, 3, 10, 30 and 100. Think about the output, sequence, and body before you start writing the loop.
```{r}
 # number to draw
    n <- 10
    # values of the mean
    mu <- c(-10, 0, 10, 100)
    normals <- vector("list", length(mu))
    for (i in seq_along(normals)) {
      normals[[i]] <- rnorm(n, mean = mu[i])
    }
    normals

# Generating a random sample from a Poisson distribution with lambda=3
set.seed(124)
pois <- rpois(100, lambda = 3)
pois[1:10]
```

2. Imagine you have a directory full of CSV files that you want to read in. files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write a for loop that will load them into a single data frame (you do not need to run anything just write code)

```{r}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
files

# Since, the number of files is known, pre-allocate a list with a length equal to the number of files.
df_list <- vector("list", length(files))
for (i in seq_along(files)) {
  df_list[[i]] <- read_csv(files[[i]])
}

print(df_list)
df <- bind_rows(df_list)
```


3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example show_mean(iris) would print:
```{r}
show_mean(iris)
```

```{r}
show_mean <- function(df, digits = 2) {
  # Get max length of all variable names in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(
        str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
          format(mean(df[[nm]]), digits = digits, nsmall = digits),
          sep = " "
        ),
        "\n"
      )
    }
  }
}
```

4. Write code that uses one of the map function to:
  a. Compute the mean of every column in mtcars
```{r}
map_dbl(mtcars, mean)
```

  b. Compute the number of unique values in each column in mpg
```{r}
map_chr(mpg, typeof)

map_int(iris, n_distinct)
# map_dbl(iris, n_distinct)
```

  c. Generate 10 random poissons (rpois) for each $\lambda$ = 1, 3, 10, 30 and 100
```{r}
map(c(-10, 0, 10, 100), ~rnorm(n = 10, mean = .))
```

5. Repeat question 2 using the map function. (The csv one)

```{r}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
files

# Since, the number of files is known, pre-allocate a list with a length equal to the number of files.
df_list <- vector("list", length(files))
map_dbl(df_list, read_csv)

print(df_list)
df <- bind_rows(df_list)
```

6. What happens when we use the map functions on vectors that aren't lists. Use the following vector for the next section:
```{r}
five_squares <- (1:5)^2
```
  a. Describe the output of using on a list map(list(five_squares),rnorm). Explain why the output turns out this way.
```{r}
map(list(five_squares),rnorm)
```

  b. What does map(five_squares,rnorm) do? Why?
```{r}
map(five_squares,rnorm)
```

  c. What does map(five_squares, rnorm, n = 5) do? Why?
```{r}
map(five_squares, rnorm, n = 5)
```





