---
title: "Skills Problem Set VI"
author: "Earnest Salgado"
date: "25/05/2021"
output: 
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
urlcolor: blue
always_allow_html: true
---
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

<!-- .Rmd files use  markdown, a text mark up language, to provide formating.--> 
<!--Text include within these strange arrows are comments and will not show up when you knit-->

This submission is my work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **ES**

Add your collaborators: **GATW**

Late coins used this pset: 0. Late coins left: 6. 
<!--You may use up to two for a given assignment.)-->

# Functions (15 points)
<!--(Notice the use of two `##`)-->

1. Write a function that transforms a vector c("a", "b", "c") into a string, a, b, and c (including the and!!). Think carefully about what the function should do if given a vector of length 1, 2, or 3.

```{r}
str_commasep <- function(x, delim = ",") {
  n <- length(x)
  if (n == 0) {
    ""
  } else if (n == 1) {
    x
  } else if (n == 2) {
    str_c(x[[1]], "and", x[[2]], sep = " ")
  } else {
    not_last <- str_c(x[seq_len(n - 1)], delim)
    last <- str_c("and", x[[n]], sep = " ")
    str_c(c(not_last, last), collapse = " ")
  }
}
```

We can test that our function is doing what we want with various vectors of different lengths:
```{r}
str_commasep("")
str_commasep("a")
str_commasep(c("a", "b"))
str_commasep(c("a", "b", "c"))
str_commasep(c("a", "b", "c", "d"))
str_commasep(c("a", "b", "c", "d", "e"))
```
2. Write a function that given your birthday (as a date), returns how old you are in years

```{r}
age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1991-03-17"))
```
3. Statistical functions
  a. Write a function to calculate the variance of a numeric vector
  
First I create a numeric vector x and then write my function that serves as the variance formula. Finally, we can check our 'variance' function against the calculated value from using var()
```{r}
y <- c(3, 6, 9, 17, 19, 25, 6)

variance <- function(x, na.rm = TRUE) {
  n <- length(x)
  m <- mean(x, na.rm = TRUE)
  sq_err <- (x - m)^2
  sum(sq_err) / (n - 1)
}

var(y)
variance(y)
```
  b. Write a function to calculate the skewness of a numeric vector
```{r}
skew <- c(1, 6, 19, 263)
skewness <- function(x, na.rm = FALSE) {
  n <- length(x)
  m <- mean(x, na.rm = na.rm)
  v <- var(x, na.rm = na.rm)
  (sum((x - m) ^ 3) / (n - 2)) / v ^ (3 / 2)
}

skewness(skew)
```
Since there are multiple definitions for skewness, we could alternatively calculate it with this function:
```{r}
skewness2 <- function(x) {
    n <- length(x)
    mean_x <- mean(x)
    sd_x <- sqrt(sum((x - mean_x)^2) / (n))
    z <- (x - mean_x) / sd_x

skewness2 <- sum(z^3) / n
skewness2
}
print(skewness2(skew))
```
  c. Use summarize_if() to calculate the mean, variance, and skewness of all numeric columns in the diamond dataset. Then, tidy the table so we have one row for each variable. (Hint: summarize_if() takes a boolean in the first position and a named vector of functions in the second position.)

```{r}
summarizeif <- function(x)  {
      diamond_mean <- summarise_if(x, is.numeric, mean, na.rm=TRUE)
      diamond_var <- summarise_if(x, is.numeric, variance, na.rm=TRUE)
      diamond_skw <- summarise_if(x, is.numeric, skewness, na.rm=TRUE)
      rbind(diamond_mean, diamond_var, diamond_skw)
}
summarizeif(diamonds)

```

4. Rename the following functions to better reflect their purpose

The function 'f' identifies whether each element of the character vector `nchar`
starts with a common string, in this case it is `prefix`. You could rename this as has_prefix. The function 'g' drops the last element, so a better name could be drop_last.
```{r}
f <- function(string, prefix){
  str_sub(string, 1, nchar(prefix)) == prefix
}

g <- function(x){
  if (length(x) <= 1)return(NULL)
  x[-length(x)]
}

```
Examples:
```{r}
has_prefix <- f
drop_last <- g

has_prefix(c("morning", "sun", "moon", "yellow"), "mo")
drop_last(c(0, 1, 2, 3, 5))
```
5. Write a greeting function that says "good morning", "good afternoon", or "good evening", depending on the time of day. (Hint: use a time argument that defaults to lubridate::now(). That will make it easier to test your function.)

```{r}
greetings=function(time=lubridate::now())
  {
}


greeting <- function(time_now = lubridate::now()) {
  hour_now <- lubridate::hour(time_now)
  
  if (hour_now < 12) {
    "good morning"
  } else if (hour_now < 18) {
    "good afternoon"
  } else {
    "good night"
  }
}
greeting()

```

# For Loops (35 points)

1. Write for loops to
  a. Compute the mean of every column in mtcars
```{r}
meancols_mtcars <- vector("double", ncol(mtcars))
    names(meancols_mtcars) <- names(mtcars)
    for (i in names(mtcars)) {
      meancols_mtcars[i] <- mean(mtcars[[i]])
    }
    meancols_mtcars
```

  b. Compute the number of unique values in each column of mpg
```{r}
 data("mpg")
    mpg_uniq <- vector("double", ncol(mpg))
    names(mpg_uniq) <- names(mpg)
    for (i in names(mpg)) {
      mpg_uniq[i] <- n_distinct(mpg[[i]])
    }
    mpg_uniq
```
  c. Generate 10 random points distributed poissons (rpois) for each $\lambda$ = 1, 3, 10, 30 and 100. Think about the output, sequence, and body before you start writing the loop.
```{r}
lambda_vector <- c(1, 3, 10, 30, 100)

rpois_fct <- matrix("double", nrow = 10, ncol = 5)
for (i in seq_along(lambda_vector)) {
  rpois_fct[,i] <- rpois(10, lambda = lambda_vector[[i]])
}
rpois_fct

```
2. Imagine you have a directory full of CSV files that you want to read in. files <- dir("data/", pattern = "\\.csv$", full.names = TRUE), and now want to read each one with read_csv(). Write a for loop that will load them into a single data frame (you do not need to run anything just write code)

```{r}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
files

df_list <- vector("list", length(files))
for (i in seq_along(files)) {
  df_list[[i]] <- read_csv(files[[i]])
}

print(df_list)
df <- bind_rows(df_list)

```
3. Write a function that prints the mean of each numeric column in a data frame, along with its name. For example show_mean(iris) would print:
```{r}
# show_mean(iris)
## [1] "Sepal.Length : 5.84333333333333"
## [1] "Sepal.Width : 3.05733333333333"
## [1] "Petal.Length : 3.758"
## [1] "Petal.Width : 1.19933333333333"
```
```{r}
show_mean <- function(df, digits = 2) {

  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(
        str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
          format(mean(df[[nm]]), digits = digits, nsmall = digits),
          sep = " "
        ),
        "\n"
      )
    }
  }
}

show_mean(iris)
```
```{r}
names_mean <- function(df) {
  df2 <- (select_if(df, is.numeric))
  
  output <- vector("double", ncol(df2)) 
  
  for (i in seq_along(df2)){
    v_names <- names(df2[,i])
    v_mean <- mean(df2[[i]], na.rm = TRUE)
    output [[i]]<- str_c(v_names, v_mean, sep = " : ") 
  }
  output
}

(names_mean(mpg))

```
4. Write code that uses one of the map function to:
  a. Compute the mean of every column in mtcars
```{r}
map_dbl(mtcars, mean)
```

  b. Compute the number of unique values in each column in mpg

Out of curiosity I pulled the types of each variable. I believe we can also use map_dbl(iris, n_distinct) to return number of unique values.
```{r}
map_chr(mpg, typeof)

map_int(iris, n_distinct)
```

  c. Generate 10 random poissons (rpois) for each $\lambda$ = 1, 3, 10, 30 and 100
```{r}
v_pois <- map(mtcars, ~ rpois(10, lambda = 100))
v_pois
```

5. Repeat question 2 using the map function. (The csv one)

```{r}
files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)
files

read_files <- map_dbl(files, ~read.csv)
```
6. What happens when we use the map functions on vectors that aren't lists. Use the following vector for the next section:
```{r}
five_squares <- (1:5)^2
five_squares
```
  a. Describe the output of using on a list map(list(five_squares),rnorm). Explain why the output turns out this way.

The output is this way because five_squares is no longer being read as a vector, but as a list. This means the function rnorm will only be run 5 times, through the list of objects within five_squares.
```{r}
map(list(five_squares),rnorm)
```
  b. What does map(five_squares,rnorm) do? Why?

This output computes normal distributions for the vector five_squares. As it generates, it takes each object as its parameter. So at five_squares[[1]], it is only 1 so the list is one value long. It continues until finally at five_squares[[5]], it identifies n = 25 so that particular list is 25 values long.
```{r}
map(five_squares,rnorm)
```
  c. What does map(five_squares, rnorm, n = 5) do? Why?

This output is the list of vectors rnorm(five_squares[[1]], n=5), rnorm(five_squares[[2]], n=5), rnorm(five_squares[[3]], n=5), rnorm(five_squares[[4]], n=5), rnorm(five_squares[[5]], n=5). It will perform the operation rnorm at each object in the vector, in our case 1 through 5, thus why its returning an output n = 5 times at each object.
```{r}
map(five_squares, rnorm, n = 5)
```





